---
title: Color Palettes in ggplot2
author: Jeff
date: '2017-11-08'
slug: color-palettes-in-ggplot2
categories:
  - R
  - Visualization
tags:
  - ggplot2
  - Visualization
  - Data Viz
  - R Markdown
draft: yes
---
```{r global_options, include=FALSE}
knitr::opts_chunk$set(fig.width=8, fig.height=7, warning=FALSE, message=FALSE)
```
ggplot2 is one of the most popular visualization packages in R for good reason.  There are several books dedicated just to this package and its learning curve can be intimidating.

Rather than attempt (foolhardily!) to give an in-depth tutorial to ggplot2, this post will show:

* one method of changing the **colour palette** *(remember that spelling for ggplot2, friends)*
* one example of changing the theme color to better suit the chosen palette.

These examples utilize NMHSA data (found [here](https://datafiles.samhsa.gov/study-dataset/national-mental-health-services-survey-2015-n-mhss-2015-ds0001-nid17098)) to show the density of VA hospitals per state/region in ever 1,000 miles.

Here’s a ggplot2 bars chart using the default palette options:
```{r load-libraries, echo=FALSE,warning=FALSE,message=FALSE}
invisible(library(ggplot2))
invisible(library(ggthemes))
invisible(theme(plot.title = element_text(hjust = 0.5)))
invisible(theme_update(plot.title = element_text(hjust = 0.5)))
size_and_count <- read.csv("~/itsdotcom/bluestnight/content/data_files/size_and_count.csv",header=TRUE)
```

```{r default-palette, echo=TRUE}
ggplot(size_and_count,aes(x=StateName, y=Density,fill=Region))+
  geom_bar(aes(reorder(StateName,Density)), stat="identity")+
  ggtitle("Number of VA Hospitals Per 1,000 Miles By State ")+
  xlab("State")+ylab("Number of VA Hospitals Per 1,000 Miles")+
  coord_flip()
```

Hey, not bad!

We’ll be using the **scale_fill_brewer** option to change our palette, an OOTB feature of ggplot2 which supports scales from colorbrewer.org.

The following examples use different types of palettes.

Here’s an example of a *diverging* palette:

```{r diverging-palette, echo=TRUE}
ggplot(size_and_count,aes(x=StateName, y=Density,fill=Region))+
  geom_bar(aes(reorder(StateName,Density)), stat="identity")+
  ggtitle("Number of VA Hospitals Per 1,000 Miles By State ")+
  xlab("State")+ylab("Number of VA Hospitals Per 1,000 Miles")+
  scale_fill_brewer(palette="Spectral")+
  coord_flip()
```

This is a *qualitative* palette:

```{r qualitative-palette, echo=TRUE}
ggplot(size_and_count,aes(x=StateName, y=Density,fill=Region))+
  geom_bar(aes(reorder(StateName,Density)), stat="identity")+
  ggtitle("Number of VA Hospitals Per 1,000 Miles By State ")+
  xlab("State")+ylab("Number of VA Hospitals Per 1,000 Miles")+
  scale_fill_brewer(palette="Pastel1")+
  coord_flip()
```

For the last example, I quickly realized the colors didn’t pop well on the gray plot area and white background. That first led me to tinker with the **ggthemes** package, then I further changed the axis colors to fit my canvas here.

Here is an example of a  sequential palette, with the above modifications:

```{r sequential-palette, echo=TRUE}
theme_update(axis.text.x = element_text(colour="green"),axis.text.y = element_text(colour="green"))
ggplot(size_and_count,aes(x=StateName, y=Density,fill=Region))+
  geom_bar(aes(reorder(StateName,Density)), stat="identity")+
  ggtitle("Number of VA Hospitals Per 1,000 Miles By State ")+
  xlab("State")+ylab("Number of VA Hospitals Per 1,000 Miles")+
  scale_fill_brewer(palette="PuBuGn")+
  theme_hc(bgcolor="darkunica")+
  coord_flip()
```

That's all there is to it! Granted, ggplot2 is incredibly customizable, but today we saw a quick way to easily change the color palette and/or theme to bring just a little something extra. See below for some reference links to dive in deeper in to colors and ggthemes with ggplot2.  

References:  
  
[ggplot2 - bars charts](http://ggplot2.tidyverse.org/reference/geom_bar.html)  
  
[ggplot2 - colour scales from colorbrewer.org](http://ggplot2.tidyverse.org/reference/scale_brewer.html)  
  
[Introduction to ggthemes](https://cran.r-project.org/web/packages/ggthemes/vignettes/ggthemes.html)  